---
title: "Getting Started with the Data API: \nSurvival Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting-started-survival-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
NOT_CRAN <- !is.null(Sys.getenv("GECO_API_PASSWORD")) && Sys.getenv("GECO_API_PASSWORD") != ''
knitr::opts_chunk$set(
  purl = NOT_CRAN,
  collapse = TRUE,
  comment = "#>",
  fig.height = 6,
  fig.width = 6
)
```

```{r setup}
library(ggplot2)
library(tidyverse)
library(rgeco)
```

First we will login to the Generable API with our username and password (here, stored in environment variables).

```{r login, eval = NOT_CRAN}
login(user = Sys.getenv('GECO_API_USER'), password = Sys.getenv('GECO_API_PASSWORD'))
demo_project <- Sys.getenv('GECO_API_TEST_PROJECT')
```

## Loading data into memory 

For this analysis we will work with survival data. The information we need is stored in the `subjects` table.

```{r, eval = NOT_CRAN}
subjects <- fetch_subjects(project = demo_project, event_type = 'overall_survival')
```

This command returns a data.frame of records, one per subject

```{r, eval = NOT_CRAN}
glimpse(subjects)
```

Notice here that we have loaded subjects + events data in one step. For endpoints that are non-terminal or that have more than one event per subject, we would want to load those data separately by calling `fetch_subjects(..., event_type = NULL)` and `fetch_events(...)`. For this project, it is reasonable to load data in one step.

This project contains synthetic data for 5 trials.

```{r, eval = NOT_CRAN}
subjects %>%
  dplyr::filter(is.na(parent_trial_id)) %>%
  distinct(trial_name)
```

Note that, in this project, we have a number of so-called "validation" trials used for testing in the context of cross-validation. For now we are going to ignore these. 

```{r, eval = NOT_CRAN}
subjects <- subjects %>%
  dplyr::filter(is.na(parent_trial_id))
```

## Compute Kaplan-Meier survival estimates

Next we will compute the Kaplan-Meier survival estimates for our 5 studies.

```{r, eval = NOT_CRAN}
km_data <- prep_km_data(data = subjects,
                        formula = survival::Surv(time = overall_survival_event_trial_day,
                                                 event = overall_survival_event_flag == 'occurred') ~
                          trial_name + trial_arm_name)
```

This function returns data in a structure that is designed to make it easy to prepare survival plots using standard ggplot2 or other libraries, and eventually to overlay those with posterior inferences from our Inference API.

```{r, eval = NOT_CRAN}
glimpse(km_data)
```

For example, we can easily plot K-M estimates using ggplot2:

```{r plot-survival, eval = NOT_CRAN}
ggplot(km_data, aes(x = time, y = estimate, colour = trial_arm_name)) + 
  facet_wrap(~ trial_name, ncol = 2) + 
  geom_step() +
  theme_minimal() + 
  ggtitle('Kaplan-Meier estimates for overall survival') +
  scale_y_continuous(labels = scales::percent) +
  guides(colour = guide_legend('Trial arm'))
```


